\section{Acronyms}\label{acronyms}

\subsection{A}\label{a}

\textbf{AMI} - Advanced Metering Infrastructure\\
Smart meter systems that collect detailed consumption data at frequent intervals (typically 15 minutes or hourly).

\textbf{API} - Application Programming Interface\\
A set of protocols and tools for building software applications that enable different systems to communicate.

\textbf{ARIMA} - AutoRegressive Integrated Moving Average\\
A statistical model for time series forecasting.

\subsection{C}\label{c}

\textbf{CCPA} - California Consumer Privacy Act\\
State privacy law regulating how businesses handle personal information.

\textbf{CICIDS2017} - Canadian Institute for Cybersecurity Intrusion Detection Dataset 2017\\
A benchmark dataset for cybersecurity research.

\subsection{D}\label{d}

\textbf{DER} - Distributed Energy Resources\\
Small-scale power generation or storage located near consumption, such as rooftop solar or batteries.

\textbf{DGA} - Dissolved Gas Analysis\\
A technique for monitoring transformer health by analyzing gases dissolved in transformer oil.

\subsection{E}\label{e}

\textbf{EAM} - Enterprise Asset Management\\
Software systems for tracking and managing physical infrastructure assets.

\textbf{ETL} - Extract, Transform, Load\\
The process of moving data from source systems to analytics platforms.

\subsection{F}\label{f}

\textbf{FERC} - Federal Energy Regulatory Commission\\
U.S. federal agency that regulates interstate electricity transmission and wholesale markets.

\subsection{G}\label{g}

\textbf{GDPR} - General Data Protection Regulation\\
European Union data privacy regulation.

\textbf{GIS} - Geographic Information System\\
Systems for capturing, storing, and analyzing geographic and spatial data.

\textbf{GHI} - Global Horizontal Irradiance\\
Total solar radiation received on a horizontal surface, used in solar forecasting.

\subsection{I}\label{i}

\textbf{IT} - Information Technology\\
Computer systems, software, and data management infrastructure.

\subsection{K}\label{k}

\textbf{KPI} - Key Performance Indicator\\
Metrics used to evaluate operational performance and success.

\subsection{L}\label{l}

\textbf{LLM} - Large Language Model\\
Advanced AI models trained on vast text corpora, capable of understanding and generating human language.

\textbf{LSTM} - Long Short-Term Memory\\
A type of recurrent neural network for time series modeling.

\subsection{M}\label{m}

\textbf{MAE} - Mean Absolute Error\\
A forecast accuracy metric measuring average prediction error magnitude.

\textbf{MAPE} - Mean Absolute Percentage Error\\
A forecast accuracy metric expressing error as a percentage of actual values.

\textbf{ML} - Machine Learning\\
A subset of artificial intelligence that enables systems to learn from data without explicit programming.

\textbf{MLOps} - Machine Learning Operations\\
Practices and tools for deploying and maintaining ML models in production.

\textbf{MSE} - Mean Squared Error\\
A forecast accuracy metric that penalizes larger errors more heavily.

\subsection{N}\label{n}

\textbf{NERC} - North American Electric Reliability Corporation\\
Organization that develops and enforces reliability standards for the North American bulk power system.

\textbf{NERC CIP} - NERC Critical Infrastructure Protection\\
Cybersecurity standards for protecting critical power infrastructure.

\textbf{NLP} - Natural Language Processing\\
AI techniques for understanding and processing human language.

\subsection{O}\label{o}

\textbf{OMS} - Outage Management System\\
Software systems that track and manage power outages and restoration efforts.

\textbf{OT} - Operational Technology\\
Hardware and software that monitors and controls physical processes, such as SCADA systems.

\subsection{P}\label{p}

\textbf{PMU} - Phasor Measurement Unit\\
High-speed sensors that measure grid voltage and current phasors for real-time monitoring.

\textbf{PUC} - Public Utility Commission\\
State regulatory agencies that oversee utility operations and rates.

\textbf{PV} - Photovoltaic\\
Solar panels that convert sunlight directly into electricity.

\textbf{PVLib} - Photovoltaic Library\\
Python package for modeling solar PV system performance.

\subsection{R}\label{r}

\textbf{RL} - Reinforcement Learning\\
A machine learning paradigm where agents learn through trial and error.

\textbf{RMSE} - Root Mean Squared Error\\
A forecast accuracy metric, the square root of mean squared error.

\textbf{ROC} - Receiver Operating Characteristic\\
A curve showing the trade-off between true positive rate and false positive rate for classification models.

\textbf{ROI} - Return on Investment\\
A measure of the financial return from an investment relative to its cost.

\subsection{S}\label{s}

\textbf{SAIDI} - System Average Interruption Duration Index\\
A reliability metric measuring the average duration of outages per customer.

\textbf{SAIFI} - System Average Interruption Frequency Index\\
A reliability metric measuring the average number of outages per customer.

\textbf{SARIMA} - Seasonal AutoRegressive Integrated Moving Average\\
An extension of ARIMA that models seasonal patterns.

\textbf{SCADA} - Supervisory Control and Data Acquisition\\
Real-time monitoring and control systems for grid equipment.

\textbf{SHAP} - SHapley Additive exPlanations\\
A method for explaining individual model predictions.

\textbf{SIEM} - Security Information and Event Management\\
Systems that collect and analyze security logs for threat detection.

\subsection{T}\label{t}

\textbf{TF-IDF} - Term Frequency-Inverse Document Frequency\\
A text feature extraction method that weights words by their importance in documents.

\subsection{Y}\label{y}

\textbf{YOLO} - You Only Look Once\\
A real-time object detection algorithm used in computer vision applications.
